---
title: "View the data structure"
format: 
  html: 
    code-fold: true
---

## Overview

```{r}
#| label: init
#| message: false
library(tidyverse)
library(targets)
tar_source()
tar_load(c(exp1_data, exp2_data, exp1_data_agg, exp2_data_agg))
```

The datafiles for the two experiments are in objects `exp1_data` and `exp2_data`. Let's first look at the structure of the data for experiment 1.

```{r}
#| label: exp1_data
head(exp1_data)
```

From Eda I know that the condition column is coded like this:

| condition | LTM      | ISI   |
|-----------|----------|-------|
| 1         | chunk    | short |
| 2         | no-chunk | short |
| 3         | chunk    | long  |
| 4         | no-chunk | long  |

Let's confirm this:

```{r}
exp1_data |>
  select(condition, chunk, gap) |>
  unique() |>
  arrange(condition)
```

there are this many trials per participant:

```{r}
max(as.numeric(exp1_data$trial)) + 1
```

## Aggregate data

This is what the aggregated data looks like:

```{r}
exp1_data_agg
```
