---
title: "Modeling with the approach reported in the draft"
format: html
---

```{r}
#| label: init
#| message: false
library(tidyverse)
library(targets)
tar_source()
tar_load_everything()
```

## Ignoring first chunk in the optimiziation

### Basic estimation of Exp1

Let's apply the modeling approach reported in the paper. We ignore the first chunk (SP1-3) while evaluating the likelihood. Eda did this because the model as implemented predicts the same performance for known and random chunks.

```{r}
#| label: model 1
start <- paper_params()
(est <- estimate_model(start, data = exp1_data_agg, exclude_sp1 = TRUE))
exp1_data_agg$pred <- predict(est, exp1_data_agg, group_by = c("chunk", "gap"))
exp1_data_agg |>
  ggplot(aes(x = gap, y = p_correct, color = chunk)) +
  geom_point() +
  geom_line() +
  geom_line(aes(y = pred), linetype = "dashed") +
  scale_color_discrete("First chunk LTM?") +
  facet_wrap(~itemtype)
```

I get startlingly different paramemter estimates. Much lower `prop` and `rate` and `tau`, higher `gain`.

### Trying different starting values

I've run this with many different starting values. We tend to end up in different regions of the parameter space (the top result close to the paper's estimates):

```{r}
fits1 |>
  filter(priors_scenario == "", exclude_sp1 == TRUE, exp == 1, deviance <= 50, convergence == 0) |>
  select(prop:convergence) |>
  arrange(gain) |>
  mutate_all(round, 3) |>
  print(n = 100)
```

### Trying priors of the gain parameter

One way to deal with that is to put a prior on the gain parameter to keep it near 25. I know priors are usually a bayesian thing, but they work with ML optimization just as well. On the next set of simulations, I used a Normal(25, 0.1) prior on the gain parameter (could have also fixed it to this value, but this gives me mroe control).

```{r}
fits1 |>
  filter(priors_scenario == "gain", exclude_sp1 == TRUE, exp == 1, deviance <= 50, convergence == 0) |>
  mutate(deviance = map2_dbl(fit, exclude_sp1, function(x, y) {
    overall_deviance(x$par, exp1_data_agg, exclude_sp1 = y)
  })) |>
  select(prop:convergence) |>
  arrange(gain) |>
  mutate_all(round, 3) |>
  print(n = 100)
```

So we do get at least some parameters that are close to that reported in the paper. The predictions with those parameters:

```{r}
fit <- fits1 |>
  filter(priors_scenario == "gain", exclude_sp1 == TRUE, exp == 1, convergence == 0) |>
  arrange(deviance) |>
  pluck("fit", 1)

exp1_data_agg$pred <- predict(fit, exp1_data_agg, group_by = c("chunk", "gap"))

exp1_data_agg |>
  ggplot(aes(x = gap, y = p_correct, color = chunk)) +
  geom_point() +
  geom_line() +
  geom_line(aes(y = pred), linetype = "dashed") +
  scale_color_discrete("First chunk LTM?") +
  facet_wrap(~itemtype)
```

### Trying priors on the rate parameter

Maybe there is a region with a higher rate that we have not explored? Let's try a prior on the rate parameter, ~ Normal(0.1, 0.01).

```{r}
fits1 |>
  filter(priors_scenario == "rate", exclude_sp1 == TRUE, exp == 1, convergence == 0) |>
  mutate(deviance = map2_dbl(fit, exclude_sp1, function(x, y) {
    overall_deviance(x$par, exp1_data_agg, exclude_sp1 = y)
  })) |>
  select(prop:convergence) |>
  arrange(deviance) |>
  mutate_all(round, 3) |>
  print(n = 100)
```

Deviance is quite much higher. Predictions?

```{r}
fit <- fits1 |>
  filter(priors_scenario == "rate", exclude_sp1 == TRUE, exp == 1, convergence == 0) |>
  mutate(deviance = map2_dbl(fit, exclude_sp1, function(x, y) {
    overall_deviance(x$par, exp1_data_agg, exclude_sp1 = y)
  })) |>
  arrange(deviance) |>
  pluck("fit", 1)

fit$par

exp1_data_agg$pred <- predict(fit, exp1_data_agg, group_by = c("chunk", "gap"))

exp1_data_agg |>
  ggplot(aes(x = gap, y = p_correct, color = chunk)) +
  geom_point() +
  geom_line() +
  geom_line(aes(y = pred), linetype = "dashed") +
  scale_color_discrete("First chunk LTM?") +
  facet_wrap(~itemtype)
```

Geater mismatch. Let's include error bars of the data:

```{r}
exp1_data |>
  group_by(id, chunk, gap, itemtype) |>
  summarise(
    n_total = dplyr::n(),
    n_correct = sum(cor),
    p_correct = mean(cor)
  ) |>
  ungroup() |>
  left_join(
    select(exp1_data_agg, chunk, gap, itemtype, pred),
    by = c("chunk", "gap", "itemtype")
  ) |>
  ggplot(aes(x = gap, y = p_correct, color = chunk)) +
  stat_summary() +
  stat_summary(aes(y = pred), linetype = "dashed", geom = "line") +
  scale_color_discrete("First chunk LTM?") +
  facet_wrap(~itemtype)
```

Paraneters seem consistent with the data (see [my notes](../notes.md)).

## Including the first chunk in the optimization

The reports above followed the approach in the current draft and excluded the first chunk from the calculation of the likelihood when optimizing the parameters. Let's include it:

```{r}
start <- paper_params()
(est <- estimate_model(start, data = exp1_data_agg, exclude_sp1 = FALSE))
exp1_data_agg$pred <- predict(est, exp1_data_agg, group_by = c("chunk", "gap"))
exp1_data_agg |>
  ggplot(aes(x = gap, y = p_correct, color = chunk)) +
  geom_point() +
  geom_line() +
  geom_line(aes(y = pred), linetype = "dashed") +
  scale_color_discrete("First chunk LTM?") +
  facet_wrap(~itemtype)
```

In this case I didn't have to use many starting values - the result is reached from almost everywhere:

```{r}
fits1 |>
  filter(priors_scenario == "", exclude_sp1 == FALSE, exp == 1, convergence == 0) |>
  arrange(deviance) |>
  select(prop:convergence) |>
  print(n = 100)
```


### With prior on rate

```{r}
fits1 |>
  filter(priors_scenario == "rate", exclude_sp1 == FALSE, exp == 1, convergence == 0) |>
  mutate(deviance = map2_dbl(fit, exclude_sp1, function(x, y) {
    overall_deviance(x$par, exp1_data_agg, exclude_sp1 = y)
  })) |>
  select(prop:convergence) |>
  arrange(deviance) |>
  mutate_all(round, 3) |>
  print(n = 100)

fit <- fits1 |>
  filter(priors_scenario == "rate", exclude_sp1 == FALSE, exp == 1, convergence == 0) |>
  arrange(deviance) |>
  pluck("fit", 1)

exp1_data_agg$pred <- predict(fit, exp1_data_agg, group_by = c("chunk", "gap"))

exp1_data_agg |>
  ggplot(aes(x = gap, y = p_correct, color = chunk)) +
  geom_point() +
  geom_line() +
  geom_line(aes(y = pred), linetype = "dashed") +
  scale_color_discrete("First chunk LTM?") +
  facet_wrap(~itemtype)
```

## Experiment 2


```{r}
start <- paper_params(exp = 2)
(est <- estimate_model(start, data = exp2_data_agg, exclude_sp1 = TRUE))
exp2_data_agg$pred <- predict(est, exp2_data_agg, group_by = c("chunk", "gap"))
exp2_data_agg |>
  ggplot(aes(x = gap, y = p_correct, color = chunk)) +
  geom_point() +
  geom_line() +
  geom_line(aes(y = pred), linetype = "dashed") +
  scale_color_discrete("First chunk LTM?") +
  facet_wrap(~itemtype)
```
